---
title: "Personal Activity Monitoring Analysis"
author: "Wayne Tipton"
date: "October 13, 2017"
output:
  md_document:
    variant: markdown_github

---
# Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. 

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

# Read in Data
The data for this assignment can be downloaded from the course web site:

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(data.table)
library(Hmisc)
setwd("C:/Users/Wayne Office Laptop/Documents/GitHub/personal-Activity-Monitoring---Peer-Graded-Assignment-")
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL,destfile = "./activity.zip")
unzip(zipfile ="./activity.zip")
pathdata <- "./"
list.files("./")
# read data
activityData <- fread("activity.csv",na.strings = c("NA"))

```


# 1. Transform and Explore Data
#### Remove missing data, change variable attributes.


```{r }
# data before transform
str(activityData)
actDataNoNa <- activityData %>%
        drop_na(steps)
#make date from character to a date type
actDataNoNa$date <- as.Date(actDataNoNa$date, "%Y-%m-%d")
str(actDataNoNa)

```

#### What is mean total number of steps taken per day ingnoring missing values?

```{r mean total steps, echo=TRUE, message=FALSE, warning=FALSE}
#Calculate the total number of steps taken per day
actDataStepsDay <- actDataNoNa %>%
        group_by(date)%>%
        summarise(totalSteps = sum(steps))
head(actDataStepsDay)
```

#### Histogram of the total number of steps taken each day.

```{r total steps histogram, echo=TRUE}
hist(actDataStepsDay$totalSteps, col = 'Blue',xlab = "Number of Steps",
     main = "Histogram of Daily Steps Taken", ylim = c(0,40))
```

#### Calculate and report the mean and median steps per day.

```{r mean and media calc, echo=TRUE}
rawMean <- mean(actDataStepsDay$totalSteps)
rawMedian <- median(actDataStepsDay$totalSteps)
raw <- cbind(rawMean,rawMedian)
raw
```
#### What is the average daily activity pattern?
#### Make a time series plot of the 5 min intervals (x-axis), and average number of steps across all days (y-axis).

```{r,plot time series, echo=TRUE}
actDataStepsInterval <- actDataNoNa %>%
        group_by(interval)%>%
        summarise(totalSteps = sum(steps))
#plot time series of the median steps for each day
plot(actDataStepsInterval$interval,actDataStepsInterval$totalSteps, type = "l",
     col="green", ylab = "Total Steps Taken", xlab = "Time Interval",
     lwd=2, ylim = c(0,12000), main=" Total Steps per 5 min Interval")
```

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r,max number of steps, echo=TRUE, warning=FALSE,}
maxInterval <- filter(actDataStepsInterval,totalSteps==max(totalSteps))
maxInterval
```
Interval 835 has the max number of steps of 10927. 

# 2. Imputing Missing Values
There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r,echo=TRUE}
sum(is.na(activityData)==TRUE)
# percentage of missing data
mean(is.na(activityData$steps))
```
There are 2304 missing step data or about 13% of the step data is missing.

#### Devise a strategy for filling in all of the missing values in the dataset.

The strategy is to fill missing data with mean steps using the Hmisc package.

#### Create a new dataset that is equal to the original dataset but with missing data filled in.
Added a imputedMean Steps column.

```{r,echo=TRUE}
actDataImpute <- activityData
actDataImpute$imputedMeanSteps <- with(actDataImpute,impute(steps,mean))
actDataImpute$date <- as.Date(actDataImpute$date, "%Y-%m-%d")
head(actDataImpute)
```

The result is each NA was replaced with the mean value of 37.3826.

#### Make a histogram of the total number of steps taken each day.

```{r,echo=TRUE}
actDataStepsImpute<- actDataImpute %>%
        group_by(date)%>%
        summarise(totalSteps = sum(imputedMeanSteps))
head(actDataStepsImpute)
# Plot histogram
hist(actDataStepsImpute$totalSteps, col = 'Orange',xlab = "Number of Steps",
     main = "Histogram of Daily Steps Taken", ylim = c(0,40)) 
```

####  Calculate and report the mean and median of the imputed steps data.

```{r,echo=TRUE}
imputeMean <- mean(actDataStepsImpute$totalSteps)
imputeMedian <- median(actDataStepsImpute$totalSteps)
imputed <- cbind(imputeMean,imputeMedian)
imputed
```

#### What is the impact of imputing missing data?

```{r, echo=TRUE}
Differencemean <- rawMean - imputeMean
DifferenceMedian <- abs(rawMedian-imputeMedian)
Difference  <- cbind(Differencemean,DifferenceMedian)
raw_imputed <- as.data.frame(rbind(raw,imputed,Difference))

colnames(raw_imputed) <- c("Mean", 'Median')
rownames(raw_imputed) <- c("Raw Data","Imputed Data", "Difference")

raw_imputed
```

The mean between imputed data, and data where NAs are ignored remained the same, while the median value increased 1.2 steps.

# Comparing Weekday and Weekend Activity.
Are there differences in activity patterns between weekdays and weekends?

#### Create a new factor variable in the dataset with two levels"weekday" and "weekend" and 

```{r,echo=TRUE}
wdays<-c('Monday','Tuesday','Wednesday','Thursday','Friday')
imputedActivityDataStepsInt <- actDataImpute 
imputedActivityDataStepsInt$dayofweek <- 
        factor((weekdays(imputedActivityDataStepsInt$date)%in%wdays),
               levels = c(TRUE,FALSE), labels = c('Weekday','Weekend'))
head(imputedActivityDataStepsInt)

meanInterval <- aggregate(imputedMeanSteps ~ interval + dayofweek, 
                          imputedActivityDataStepsInt, FUN = mean)
head(meanInterval)

#create weekday and weekend step data views for plotting
stepsWD<-meanInterval %>%
        filter(dayofweek=='Weekday')

stepsWE <- meanInterval %>% 
        filter(dayofweek=='Weekend')
```

#### Create time series plots for weekday and weekend activity. 

```{r,echo=TRUE}
# plot weekdays
par(mfrow=c(2,1))
par(mar=c(2,2,1,1))
plot(stepsWD$interval,stepsWD$imputedMeanSteps, type = "l",
     lwd = 2, ylim = c(0,225), main = "Weekday Steps",
     ylab = "Number of Steps", xlab = "Interval", col = "purple")

#plot weekends
plot(stepsWE$interval,stepsWE$imputedMeanSteps, type = "l",
     lwd = 2, ylim = c(0,225), main = "Weekend Steps",
     ylab = "Number of Steps", xlab = "Interval", col = "dark green")

```

Based on the plots, there is relatively more weekend activity compared to weekday activity.














